Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-15 paper/ms.tex
update ms

{\it How do we use the bit-flags?} By default, we assume that a star
is suitable for gyrochronology if none of bits zero through nine
(inclusive) are raised.  For analyses that require a homogeneous
stellar rotation selection function, we also require bit 11 to be
zero.

\subsection{Planet Quality Flags}
\label{subsec:plflags}
Some planets are more reliably identified than others.  We use the
following additional quality indicators to assess the reliability and
utility of a planet, and assemble them into a separate bit-flag,
$Q_{\rm planet}$.

@@

Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
stars and known planet-hosts in the Kepler field.  In detail, these
``histograms'' are sampled from the posterior probability
distributions by drawing ten random samples from each posterior
distribution, and then computing the histogram of the resulting
samples.  We truncated the plots near $4$\,Gyr since this is the
cluster age out to which our interpolation-based method is calibrated.

For Sun-like stars, the fraction of rotation signals detectable by
Kepler is near unity up to $\approx$3\,Gyr, and it
drops to almost zero by $\approx$5\,Gyr \citep{2022ApJ...937...94M}.
This trend has the opposite functional form from our derived age
distributions.

To quantify the amount by which young stars are more rare than old, we
can group stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr as ``young'',
``intermediate-age'', and ``old''.  A simple counting exercise from
Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
times as many old stars in the Kepler field as young stars.
Similarly, there are \ratioobtoybplanets\ times as many old planet
hosts as young planet hosts.  Focusing only on the tails of the
distributions (0-0.3\,Gyr and 2.7-3\,Gyr), the implication is that the
formation rate of stars in the Kepler field decreased by a factor of
\ratiosfr$\pm$\uncratiosfr\ over the past three billion years.

In terms of detected planet counts, our rotation-based ages for the
Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.  
The stellar ages and planetary
properties of all of these systems are given in
Table~\ref{tab:planets}.

@@

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
most constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
Different sub-groups of the complex may vary in age by $\approx$50\%
(\citealt{Bouma_2022b}; Kerr~et al.,~in preparation).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
subgroup.

@@

%Restricting our attention to the
%$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
%the respective counts are \minageltonegyryesconsistent,
%\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
%Rephrased, in the overall sample, \fracconsistentallages\% of stars
%have consistent rotation and lithium-based ages
%(\fracpotentiallyconsistentallages\% potentially consistent).  In the
%sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
%ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
%potentially consistent).  We feel it necessary to distinguish
%``maybe'' consistent cases because of lingering uncertainties in how
%the dispersion of the lithium data evolve over the first gigayear.

\subsubsection{New Young Planets}
%
%The astrophysical processes behind rotation and lithium evolution are
%independent.
%Therefore 
%whenever the two age indicators agree within a
%narrowly-constrained range, confidence in the true age
%increases.
%The region of greatest overlap between rotation and lithium ages is
%dictated by the timescale for lithium depletion: for stars with
%temperatures between 4500-5500\,K with ages $\lesssim$1\,Gyr, we can
%hope to obtain independent two-sided constraints
%\citep{Bouma_2023,Jeffries_2023}.

Let us restrict our attention to blemish-free stars, with blemish-free
``confirmed'' planets: $Q_{\rm star}$ must not have any of bits zero
through nine raised, and $Q_{\rm planet}$ must have no quality flags
raised at all.  Under this constraint, our results include
\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
provides an age constraint while $t_{\rm Li}$ is one-sided.
%Two of the former systems are in clusters (Kepler-1643 and Kepler-1928);
%five of the latter systems are in Theia~520.

@@

\subsection{Caveats, Limitations}

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates
of main sequence stars.  Binarity is one example: even
intermediate-separation ($\sim$10\,AU) binaries are biased toward
rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
were interested in not only the overall age distribution of the Kepler
field, but also in the reliability of individual ages for young planet
hosting stars.  The quality flagging scheme in Table~\ref{tab:planets}
provides one path toward this end.  The other method is to consider
the cases for which lithium
and rotation overlap in sensitivity -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
for such stars include \ltonegyrhighqconfirmedtwosided\ planets
orbiting ``high-quality'' stars for which the two age indicators are
both constraining and in agreement (with $Q_{\rm star}$ bits zero
through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a precise
lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

{\it Field star ages from other surveys}:
While this study focused on deriving rotation and lithium-based ages
for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF, and
ATLAS have different selection functions.
Applying rotation-based age dating to their discoveries could clarify
the youth and evolutionary trends of these populations.  Future
missions like PLATO \citep{Rauer14} and Roman Galactic Plane Time
Domain Survey  will further expand the sample of stars with precisely
determined ages, including those known to host planets.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include only the youngest
  confirmed Kepler planets, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode any given $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
%  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main-sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  Bit 11: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
  The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
  not be raised, and for a planet to be ``reliabe'', we suggested
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}

\startlongtable
\begin{deluxetable*}{lllllllrrrrrr}
  \tabletypesize{\scriptsize}
  \tablecaption{
  Ages of Kepler target stars derived from rotation periods.  The
  full machine-readable table includes \nuniqstarsantosrot\ stars
  with reported rotation periods from \citetalias{Santos_2019} and
  \citetalias{Santos_2021}, \nuniqstarfinitegyroage\ of which have
  finite reported gyrochrone ages.  A random set of these stars is
  shown here to give an idea for form and content.  The quality
  bit-flag is as in Table~\ref{tab:planets}; requiring bits zero to
  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
  which gyrochronology is likely to be valid.  \label{tab:stars}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
  -- &   -- & K & days &  Myr &    int  \\
  \hline
  \input{table_star_gyro} %\\
  %\hline
  \enddata
% \tablecomments{This table includes X, Y, Z... The machine-readable version,
%   available online, includes additional columns for .... 
% }
\end{deluxetable*}

>>>>

{\it How do we use the bitmasks?} By default, we assume that a star
is suitable for gyrochronology if none of bits zero through nine
(inclusive) are raised.  For analyses that require a homogeneous
stellar rotation selection function, we also require bit 11 to be
zero.

\subsection{Planet Quality Flags}
\label{subsec:plflags}
Some planets are more reliably identified than others.  We use the
following additional quality indicators to assess the reliability and
utility of a planet, and assemble them into a separate bitmask,
$Q_{\rm planet}$.

@@

The stellar ages and planetary properties of the systems discussed in
the previous sections are given in Tables~\ref{tab:planets}
and~\ref{tab:stars}.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
age distributions of the stars and known planet-hosts.  These
``histograms'' are sampled from the posterior probability
distributions by drawing ten random samples from each posterior
distribution, and then computing the re-weighted histogram of the
resulting samples.  The uncertainties assume Poisson statistics.  We
truncated the plots near $4$\,Gyr since this is the cluster age out to
which our interpolation-based method is calibrated.
We refer to this as a ``demographic cliff'' in reference to an
impending crisis expected to befall higher education.

The age distributions of Kepler stars and planet hosts are
statistically quite similar.  A simple diagnostic of whether the two
distributions are drawn from the same underlying distribution is the
Kolmogorov-Smirnov test;  we calculated this statistic in a manner
that accounts for the Poisson uncertainties by performing 100 random
draws of 70\% of the stars from each sample.  The resulting median
$p$-values were $\approx$0.01 for both rows of
Figure~\ref{fig:hist_tgyro}.  This degree of similarity suggests that
the ``common wisdom'' that young transiting planets are hard to detect
because their host stars are photometrically variable requires some
nuance.  If true, this statement only holds for the very youngest
stars, since at the age resolution of this study, there appears to be
little to no selection bias against sub-gigayear planets.

An important separate selection effect concerns Kepler's ability to
detect the rotation signals.  \citet{2022ApJ...937...94M} studied this
issue, and found for Sun-like stars that the fraction of rotation
signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
that it drops to almost zero by $\approx$5\,Gyr .  This trend is
opposite in functional form to our derived age distributions.

We can quantify the relative counts of stars as a function of age by
labelling stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr as ``young'',
``intermediate-age'', and ``old''.  A simple counting exercise from
Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
times as many old stars in the Kepler field as young stars.
Similarly, there are \ratioobtoybplanets\ times as many old planet
hosts as young planet hosts.  Focusing only on the tails of the
distributions (0-0.3\,Gyr and 2.7-3\,Gyr), the implication is that the
formation rate of stars in the Kepler field decreased by a factor of
\ratiosfr$\pm$\uncratiosfr\ over the past three billion years.

In terms of detected planet counts, our rotation-based ages for the
Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.

@@

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
most constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
Different sub-groups of the complex may vary in age by $\approx$50\%
(\citealt{Bouma_2022b}; Kerr~et al.,~submitted).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
subgroup.

@@

\subsubsection{New Young Planets}

Let us restrict our attention to blemish-free stars, with blemish-free
``confirmed'' planets: $Q_{\rm star}$ must not have any of bits zero
through nine raised, and $Q_{\rm planet}$ must have no quality flags
raised at all.  Under this constraint, our results include
\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
provides an age constraint while $t_{\rm Li}$ is one-sided.

@@

\subsection{Caveats \& Limitations}

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than stellar age and mass can
influence stellar rotation rates.  Binarity is one example: even
intermediate-separation ($\sim$10\,AU) binaries are biased toward
rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
is that our rotation-based ages used photometric effective
temperatures (Section~\ref{sec:stellarprops}), even though
spectroscopic temperatures are available for the planet-hosting
subset.  This decision was driven by a desire for homogeneity 
irrespective of planet-hosting status.  It may however have
incurred small systematic errors in the implied ages.

In this work, we were interested in not only the overall age
distribution of the Kepler field, but also in the reliability of
individual ages for young stars known to host planets.  The quality
bitmask scheme in Table~\ref{tab:planets} provided one path toward
this end.  Our other method was to compare the cases for which lithium
and rotation overlap in sensitivity -- G and K dwarfs with
$t$$\lesssim$1\,Gyr.  Our results for such stars include
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars for which the two age indicators are both constraining and in
agreement (stars with $Q_{\rm star}$ bits zero through nine not
raised, and no planet quality flags raised).  Conversely, we found
only one system, Kepler-786, with a precise lithium age
(\kepseveneightsix\,Myr), but no rotation signal.

@@

{\it Field star ages from other surveys}:
While this study focused on deriving rotation and lithium-based ages
for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF,
and ATLAS have different selection functions.  Applying rotation-based
age dating to their discoveries could clarify the youth and
evolutionary trends of these populations.  Future prospects like PLATO
\citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
Galactic Plane Time Domain Survey seem poisted to further expand the
sample of stars with precisely determined ages, including those known
to host planets.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include only the youngest
  confirmed Kepler planets, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode any given $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
%  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main-sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  Bit 11: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
  not be raised, and for a planet to be ``reliabe'', we suggested
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}

\startlongtable
\begin{deluxetable*}{lllllllrrrrrr}
  \tabletypesize{\scriptsize}
  \tablecaption{
  Ages of Kepler target stars derived from rotation periods.  The
  full machine-readable table includes \nuniqstarsantosrot\ stars
  with reported rotation periods from \citetalias{Santos_2019} and
  \citetalias{Santos_2021}, \nuniqstarfinitegyroage\ of which have
  finite reported gyrochrone ages.  A random set of these stars is
  shown here to give an idea for form and content.  The quality
  bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
  which gyrochronology is likely to be valid.  \label{tab:stars}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
  -- &   -- & K & days &  Myr &    int  \\
  \hline
  \input{table_star_gyro} %\\
  %\hline
  \enddata
% \tablecomments{This table includes X, Y, Z... The machine-readable version,
%   available online, includes additional columns for .... 
% }
\end{deluxetable*}
